Stage-Splitting a Modal Language

### 2014-10-06 paper/semantics.tex
added remarks

Partial values are stage \bbone\ terms that have been fully
evaluated, but which may contain stage-\bbtwo\ variables wrapped in $\next$
blocks.  In the example above, 
\begin{lstlisting} 
1`(next {`2`yhat`1`},7)`
\end{lstlisting}
is a partial value.
Residuals ($\ressym$es) are \langTwo\ terms---stage \bbtwo\ terms whose
stage \bbone\ subexpressions have all been fully evaluated. In the example
above,
\begin{lstlisting} 
2`let zhat = yhat*yhat*7 in zhat`
\end{lstlisting}
is a residual,
These classifications are similar in that they both indicate that stage \bbone\ reduction is finished.
They are different in that ``partial value" is a classification of stage~\bbone\ terms,
while ``residual" is a classification of stage-\bbtwo\ terms.

@@

At $\next~e$, in contrast, $\redonesym$ must speculate into $e$ for stage~\bbone\ 
subexpressions to evaluate.  (See \ref{fig:diaSemanticsNP}.)
This is implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual. 
For the most part (\ref{fig:diaSemanticsSpec}), speculation simply recurs into
all subexpressions; at $\prev$, however, it resumes $\redonesym$ evaluation,
which produces a future context $\xi$ and (by canonical forms) an expression
$\next\{\hat y\}$. $\xi$ is then reified using $\reifysym$ into a series of let
bindings enclosing $\mathtt{\hat y}$.

Once speculation reduces $e$ to a residual $q$, the output of $\redonesym$ is a
fresh variable wrapped in a $\next$ block ($\next~\hat y$), along with a future
environment which maps that variable to the residual ($\hat y \mapsto q$).

Within speculation lies a subtle---if perhaps unintuitive---feature.  
Observe that speculation will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
statement in its search for stage~\bbone\ code. 
Thus the evaluation of that stage~\bbone\ code will occur {\em regardless of the eventual value of the predicate},
and so a term like 
\begin{lstlisting} 
next{
  2`if true 
  then `hold{1`3+4`} 
  2`else` prev{1`spin() (* loops forever *)`}
}
\end{lstlisting}
will fail to evaluate at stage \bbone.
This behavior is why the judgment is named ``speculation."
Although it may seem undesirable, this speculative behavior is a deliberate feature.
Indeed it is critical to the quickselect example, 
wherein all three branches of the $n$-$k$ comparison are evaluated at stage~\bbone.
As always, it is the programmer's responsibility to ensure that her staging annotations 
do not cause a program to loop forever.

@@

>>>>

\ur{Partial values and residuals should be ``definitons'', i.e.,
  use italics here.}

Partial values are stage \bbone\ terms that have been fully
evaluated, but which may contain stage-\bbtwo\ variables wrapped in $\next$
blocks.  In the example above, 
\begin{lstlisting} 
1`(next {`2`yhat`1`},7)`
\end{lstlisting}
is a partial value.
Residuals ($\ressym$es) are \langTwo\ terms---stage \bbtwo\ terms whose
stage \bbone\ subexpressions have all been fully evaluated. In the example
above,
\begin{lstlisting} 
2`let zhat = yhat*yhat*7 in zhat`
\end{lstlisting}
is a residual,
\ur{Comma above! ``classification'' sounds odd to me.  just say they
  ``are stage-1 terms.''}

These classifications are similar in that they both indicate that stage \bbone\ reduction is finished.
They are different in that ``partial value" is a classification of stage~\bbone\ terms,
while ``residual" is a classification of stage-\bbtwo\ terms.

\ur{I would recommend using ``1-value'' and ``2-value'' instead of
  partial values and residuals.  The former are more accurate, more
  direct, more succinct, and does not lead to confusions with other
  residuals.}

\ur{Many incorrectly hyphenated ``partial values'' in this text.}

@@

\ur{drop ``must''.}
At $\next~e$, in contrast, $\redonesym$ must speculate into $e$ for stage~\bbone\ 
subexpressions to evaluate.  (See \ref{fig:diaSemanticsNP}.)
This is implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual. 
For the most part (\ref{fig:diaSemanticsSpec}), speculation simply recurs into
all subexpressions; at $\prev$, however, it resumes $\redonesym$ evaluation,
which produces a future context $\xi$ and (by canonical forms) an expression
$\next\{\hat y\}$. $\xi$ is then reified using $\reifysym$ into a series of let
bindings enclosing $\mathtt{\hat y}$.

\ur{This is really a continuation of the previous paragraph.}
Once speculation reduces $e$ to a residual $q$, the output of $\redonesym$ is a
fresh variable wrapped in a $\next$ block ($\next~\hat y$), along with a future
environment which maps that variable to the residual ($\hat y \mapsto q$).

Within speculation lies a subtle---if perhaps unintuitive---feature.  
Observe that speculation will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
statement in its search for stage~\bbone\ code. 
Thus the evaluation of that stage~\bbone\ code will occur {\em regardless of the eventual value of the predicate},
and so a term like 
\begin{lstlisting} 
next{
  2`if true 
  then `hold{1`3+4`} 
  2`else` prev{1`spin() (* loops forever *)`}
}
\end{lstlisting}
will fail to evaluate at stage \bbone.
\ur{This might be standard in the PE literature.  If that is the case
  there is no need to elaborate as much.}
This behavior is why the judgment is named ``speculation."
Although it may seem undesirable, this speculative behavior is a deliberate feature.
Indeed it is critical to the quickselect example, 
wherein all three branches of the $n$-$k$ comparison are evaluated at stage~\bbone.
As always, it is the programmer's responsibility to ensure that her staging annotations 
do not cause a program to loop forever.

@@

\ur{The following several paragraphs should probably be somewhere
  else, further ahead, or further back when we discuss the
  relationship with PE.}

### 2014-10-06 paper/splitting.tex
added remarks

\section{Splitting Algorithm}
\label{sec:splitting}

@@

@@

There are many reasonable ways to achieve this goal, 
and we choose one that preserves as much structure as possible from $v$.
Formally, partial value splitting comprises a single judgment, $v \vsplito [v_1,v_2]$, the rules for which are given in \cref{fig:valSplit}.
It essentially acts distributively at product, sum, and recursive types, 
doing little more than recuring into subvalues and duplicating the structure to both results.
At base types, partial value splitting assigns all content ({\em i.e.} the integer or boolean) to stage \bbone, and uses a nullary tuple placeholder for stage \bbtwo.
Conversely at $\fut$ types, splitting assigns all content ({\em i.e.} the environment reference) to stage \bbtwo, and uses a nullary tuple for stage \bbone.

Functions are inherently more complicated. 
Since the body of a function is an open term, the partial value splitting rule for functions must rely on general expression splitting, which is covered momentarily.
The most we can say here is a function splits into two other functions, 
whose bodies correspond to the stage \bbone\ and stage \bbtwo\ components of the original function's body.

@@

We now consider the problem of splitting expressions in \lang, starting with stage \bbtwo\ terms.  
The goal here is to end up with two new terms: one that has all of the stage \bbone\ content of the original, and one with all of the stage \bbtwo\ content. 
These are respectively called the {\em precomputation} and {\em residual}.
\footnote{We also used the term {\em residual} for the result of partial evaluation.
Where necessary, we will distinguish between these concepts by using the terms {\em splitting-residual} and {\em evaluation-residual}.}

For example, consider $e=``\verb|hold{1+2} < hold{3+4}-5|"$.
We can split this into two terms, $p=``\verb|(1+2,3+4)|"$ and $r=``\verb|fn (L1,L2) => L1 < L2-5|"$, 
with the as yet imprecise property that $r$ applied to $p$ yields the same final value as $e$.
In general, all we're doing here is finding the contained nuggets of stage \bbone\ code and pulling them into a big tuple,
an action we call {\em lifting}.

Note that lifting and masking are different, despite both being within the purview of splitting.
They particularly differ in how the resulting stage \bbone\ parts are used.
The $v_1$ of partial value splitting is essentially the stage \bbone\ content of {\em result} of computation, 
and it's accessible to whatever surrounding piece of code refers to the value being split.
In contrast, the $p$ of stage \bbtwo\ expression splitting is only meant to support the stage \bbtwo\ part,
and is not accessible externally.

@@

The splitting algorithm comprises two judgments, $\splitonesym$ and $\splittwosym$,
which are in rule-for-rule correspondence with the typing rules $@\bbone$ and $@\bbtwo$.
For both, the input is a term in \lang, and the output is two terms in \langmono, an
unstaged language.  The grammar for \langmono\ is given in \ref{fig:monoGrammar}.  It has no
staging features (we say that implicitly it has only one stage), and it is untyped.  
Modulo these differences, \langmono\ has all of the same features as \lang, 
although this is a matter of taste
\footnote{Since \langmono\ is untyped, we could encode everything with just functions.}.

Specifically, the $\splitonesym$ judgment sends a stage \bbone\ term ($e$) to a combined term ($c$) and residual ($l.r$),
while $\splittwosym$ sends a stage \bbtwo\ term ($e$) to a precomputation ($p$) and a residual ($l.r$).
For concision, we represent residuals as a term open on a single variable, rather than as functions.
For both judgments, we use a context ($\Gamma$) to keep track of the open variables of $e$.

The rules\footnote{The rules are written using patterns, including the open variable of the residual.} 
for splitting $\next$, $\prev$, and $\pause$ are given in \ref{fig:termSplitOne}.
The rule for $\next$ simply tuples up the precomputation of its subexpression with a trivial immediate result,
while the rule for $\prev$ projects the combined result of its subexpression to just the precomputation.
The $\pause$ rule treats the entire combined result of its subexpression as a precomputation, 
and projects out the integer result in the residual\footnote{The residual of an integer expression is usually trivial, 
but we have to include it here for termination purposes.}

>>>>

\section{Splitting Algorithm}
\label{sec:splitting}
\ur{First para should probably be somewhere else concerning problem
  definition and relationship to PE.  Based on this definition, it
  might be good and correct to argue that this is more or less an
  orthogonal problem.}

@@

\ur{Reduce commentary.  Two sentences connected with an ``and.'' Split
  sentences.}

@@

\ur{Next sentence is unnecessary commentary. It also sounds of
  questinoable accuracy, not clear to me that there are many ways to
  do this.}  There are many reasonable ways to achieve this goal, and
we choose one that preserves as much structure as possible from $v$.
Formally, partial value splitting comprises a single judgment, $v
\vsplito [v_1,v_2]$, the rules for which are given in
\cref{fig:valSplit}.  It essentially acts distributively at product,
sum, and recursive types, doing little more than recuring into
subvalues and duplicating the structure to both results.  At base
types, partial value splitting assigns all content ({\em i.e.} the
integer or boolean) to stage \bbone, and uses a nullary tuple
placeholder for stage \bbtwo.  Conversely at $\fut$ types, splitting
assigns all content ({\em i.e.} the environment reference) to stage
\bbtwo, and uses a nullary tuple for stage \bbone.

\ur{The case for ``fold'' seems missing in partial value splitting rules.}

\ur{Drop, ``the most we can say.''  Be direct and explain.}
Functions are inherently more complicated.  Since the body of a
function is an open term, the partial value splitting rule for
functions must rely on general expression splitting, which is covered
momentarily.  The most we can say here is a function splits into two
other functions, whose bodies correspond to the stage \bbone\ and
stage \bbtwo\ components of the original function's body.

@@

We now consider the problem of splitting expressions in \lang, starting with stage \bbtwo\ terms.  
The goal here is to end up with two new terms: one that has all of the stage \bbone\ content of the original, and one with all of the stage \bbtwo\ content. 
These are respectively called the {\em precomputation} and {\em residual}.
\footnote{We also used the term {\em residual} for the result of
  partial evaluation.  Where necessary, we will distinguish between
  these concepts by using the terms {\em splitting-residual} and {\em
    evaluation-residual}.}  \ur{The reuse of residual is bothersome,
  we should use different terminology, 1-value and 2-value perhaps.}

\ur{We need a ``next'' I believe.  All our examples should start at
  stage-1 for consistency.}
For example, consider $e=``\verb|hold{1+2} < hold{3+4}-5|"$.
We can split this into two terms, $p=``\verb|(1+2,3+4)|"$ and $r=``\verb|fn (L1,L2) => L1 < L2-5|"$, 
with the as yet imprecise property that $r$ applied to $p$ yields the same final value as $e$.
In general, all we're doing here is finding the contained nuggets of stage \bbone\ code and pulling them into a big tuple,
an action we call {\em lifting}.

\ur{I have no idea what the next paragraph is trying to say.}
Note that lifting and masking are different, despite both being within the purview of splitting.
They particularly differ in how the resulting stage \bbone\ parts are used.
The $v_1$ of partial value splitting is essentially the stage \bbone\ content of {\em result} of computation, 
and it's accessible to whatever surrounding piece of code refers to the value being split.
In contrast, the $p$ of stage \bbtwo\ expression splitting is only meant to support the stage \bbtwo\ part,
and is not accessible externally.

@@

\ur{``for both'' below is ambiguous.}
The splitting algorithm comprises two judgments, $\splitonesym$ and $\splittwosym$,
which are in rule-for-rule correspondence with the typing rules $@\bbone$ and $@\bbtwo$.
For both, the input is a term in \lang, and the output is two terms in \langmono, an
unstaged language.  The grammar for \langmono\ is given in \ref{fig:monoGrammar}.  It has no
staging features (we say that implicitly it has only one stage), and it is untyped.  
Modulo these differences, \langmono\ has all of the same features as \lang, 
although this is a matter of taste
\footnote{Since \langmono\ is untyped, we could encode everything with just functions.}.
\ur{Drop commentary about taste and such.  Explain that we could have
  used lambda calculus as target but this is more direct.}

\ur{Drop ``specifically''.}
Specifically, the $\splitonesym$ judgment sends a stage \bbone\ term ($e$) to a combined term ($c$) and residual ($l.r$),
while $\splittwosym$ sends a stage \bbtwo\ term ($e$) to a precomputation ($p$) and a residual ($l.r$).
For concision, we represent residuals as a term open on a single variable, rather than as functions.
For both judgments, we use a context ($\Gamma$) to keep track of the
open variables of $e$.
\ur{Case (pl./si.) mismatch in ``we represent ...'' sentence above.}

\ur{Reformat rules to make them fit into the paper.}

\ur{Explain the rules.  This is the most important part of the
  paper. Work needed here.}
The rules\footnote{The rules are written using patterns, including the open variable of the residual.} 
for splitting $\next$, $\prev$, and $\pause$ are given in \ref{fig:termSplitOne}.
The rule for $\next$ simply tuples up the precomputation of its subexpression with a trivial immediate result,
while the rule for $\prev$ projects the combined result of its subexpression to just the precomputation.
The $\pause$ rule treats the entire combined result of its subexpression as a precomputation, 
and projects out the integer result in the residual\footnote{The residual of an integer expression is usually trivial, 
but we have to include it here for termination purposes.}
